\documentclass{article}
\usepackage{../math_template}

\author{Jonathan Kasongo}
\title{Mathematical Olympiad Challenges}

\begin{document}
\maketitle
\textit{Problems taken from the book by the same name as the title, by
Andreescu and Gelca.}

\begin{problem}{Chapter 1.9 --- Problem 1}
Prove that
$$
\frac{\sin x}{\cos x} + \frac{\sin 2x}{\cos^2 x} + \cdots +
\frac{\sin nx}{\cos^n x} = \cot x - \frac{\cos (n+1)x}{\sin x \cos^n x}
$$
\end{problem}

\begin{solution}{Solution}
The heart of the problem is to consider the related sequences,
$$
a_k = \frac{\sin kx}{\cos^k x}, \quad b_k = \frac{\cos kx}{\cos^k x}.
$$
We apply the cosine addition formula on $b_k$,
\[
\begin{aligned}
b_k &= \frac{\cos [(k-1)x] \cos x - \sin [(k-1)x] \sin x}{\cos^k x}\\
&= b_{k-1} - a_{k-1} \tan x\\
\end{aligned}
\]
which implies that $a_{k} \tan x = b_k - b_{k+1}$.
Indeed the LHS of our identity now telescopes down to,
\[
\begin{aligned}
a_1 + a_2 + \cdots + a_n &= (b_1 - b_{n+1})\cot x\\
&= \cot x - \frac{\cos (n+1)x}{\sin x \cos^n x}. \ \blacksquare \\
\end{aligned}
\]
\end{solution}
\newpage

\begin{problem}{Chapter 1.9 --- Problem 3}
Let $n$ be a positive integer and $a$ a real number, such that $a/\pi$ is
an irrational number. Compute
$$
\frac{1}{\cos a - \cos 3a} + \frac{1}{\cos a - \cos 5a} + \cdots +
\frac{1}{\cos a - \cos (2n+1)a}
$$
\end{problem}

\begin{solution}{Solution}
Motivated by the sum to product formulae we notice that
\[
\begin{aligned}
\cos a - \cos (2k+1)a &= \cos [(k+1)a - ka] - \cos [(k+1)a + ka] \\
&= 2\sin [(k+1)a] \sin (ka).
\end{aligned}
\]
So we wish to compute
$$
S = \sum_{k=1}^n \frac{1}{2\sin [(k+1)a] \sin(ka)}.
$$
If one looks for a function that might telescope into the general term of
this series then we may figure out that
\[
\begin{aligned}
\frac{\cos(k+1)a}{\sin(k+1)a} - \frac{\cos(ka)}{\sin(ka)} &=
\frac{\cos[(k+1)a] \sin(ka) - \cos(ka) \sin[(k+1)a]}{\sin(ka)\sin[(k+1)a]}\\
&= \frac{\sin(-a)}{\sin(ka)\sin(k+1)a}.
\end{aligned}
\]
Armed with this fact it is clear that we can re-write our sum as
$$
S = -\frac{1}{2\sin a} \sum_{k=1}^n [\cot[(k+1)a] - \cot (ka)]
$$
which telescopes down to
$$
S = \frac{1}{2\sin a}(\cot a - \cot (n+1)a). \ \blacksquare
$$
\end{solution}
\newpage

\begin{problem}{Chapter 3.2 --- Problem 2}
Show that each of the numbers
$$
\frac{107811}{3}, \frac{110778111}{3}, \frac{111077781111}{3}, \ldots
$$
is the cube of a positive integer.
\end{problem}

\begin{solution}{Solution}
First we try to characterise the pattern given in these numbers. Let the
$k$-th number in this list be $n_k$, then

$$
3n_k = \underbrace{11\ldots1}_{k \text{ times}}\ 0 \
       \underbrace{77\ldots7}_{k \text{ times}}\ 81 \
       \underbrace{11\ldots1}_{k \text{ times}}.
$$

After looking at small cases and guessing, one finds that these numbers
are of the form $(33\ldots3)^3$. The solution is to show that
$$
n_k = (\underbrace{33\ldots3}_{k+1 \text{ times}})^3.
$$
Notice that we can write (make sure to pay careful attention
to the summation bounds, it's easy to mess up)
\[
\begin{aligned}
3n_k &= \sum_{r = 2k+3}^{3k+2} (10^r) + 7\sum_{r = k+2}^{2k+1} (10^r) +
81(10^{k}) + \sum_{r = 0}^{k-1} (10^r)\\
&= \left( 1 + 7(10^{k+2}) + 10^{2k+3} \right) \frac{10^k - 1}{9} + 81(10^k)
\\
&= -\frac19 + 30\frac{10^k}{9} - 300\frac{10^{2k}}{9} + 1000\frac{10^{3k}}{9}\\
n_k &= -\frac{1}{27} + 30\frac{10^k}{27} - 300\frac{10^{2k}}{27} +
1000\frac{10^{3k}}{27}\\
&= -\frac{1}{27} + 3\frac{10^{k+1}}{27} - 3\frac{10^{2k+2}}{27} +
\frac{10^{3k+3}}{27} \\
&= \left(\frac{10^{k+1} - 1}{3}\right)^3
\end{aligned}
\]
which is exactly the number $(\underbrace{33\ldots3}_{k+1 \text{ times}})^3$
as desired. $\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 1.9 --- Problem 5}
Compute
$$
\frac{\tan 1}{\cos 2} + \frac{\tan 2}{\cos 4} + \cdots +
\frac{\tan 2^n}{\cos 2^{n+1}}.
$$
\end{problem}

\begin{solution}{Solution}
We observe the following fact.

\[
\begin{aligned}
\tan 2^{k+1} - \tan 2^k &=
\frac{\sin 2^{k+1}}{\cos 2^{k+1}} - \frac{\sin 2^k}{\cos 2^k} \\
&=
\frac{\sin 2^{k+1} \cos 2^k - \sin 2^k \cos 2^{k+1}}{\cos 2^k \cos 2^{k+1}} \\
&= \frac{\sin 2^k}{\cos 2^k \cos 2^{k+1}}\\
&= \frac{\tan 2^k}{\cos 2^{k+1}}.
\end{aligned}
\]

Thus we can re-write the sum as
$$
\sum_{k=0}^n (\tan 2^{k+1} - \tan 2^k)
$$
which telescopes to $\tan 2^{n+1} - \tan 1$. $\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 2.3 --- Problem 3}
Evaluate
$$
\sum_{k=1}^\infty \frac{6^k}{(3^k - 2^k)(3^{k+1} - 2^{k+1})}.
$$
\end{problem}

\begin{solution}{Solution}
Define $c_k = 3^k - 2^k$. We have the recurrence
\[
\begin{aligned}
c_{k+1} &= 2c_k + 3^k\\
2^k c_{k+1} - 2^{k+1} c_k &= 6^k\\
\frac{2^k}{c_k} - \frac{2^{k+1}}{c_{k+1}} &= \frac{6^k}{c_k c_{k+1}}
\end{aligned}
\]

So the sum to evaluate is just
$$
\sum_{k=1}^\infty \frac{6^k}{c_k c_{k+1}} =
\sum_{k=1}^\infty \left( \frac{2^k}{c_k} - \frac{2^{k+1}}{c_{k+1}} \right)
$$
which telescopes down to just $2^1/c_1 - 0 = 2$. $\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 2.3 --- Problem 4}
The sequence $(x_n)_n$ is defined by $x_1 = \frac12$,
$x_{k+1} = x_k^2 + x_k$. Find the greatest integer less than
$$
\frac{1}{x_1 + 1} + \frac{1}{x_2 + 1} + \cdots + \frac{1}{x_{100} + 1}.
$$
\end{problem}

\begin{solution}{Solution}
We notice that
$$
\sum_{k=1}^{100} \frac{1}{x_k + 1} =
\sum_{k=1}^{100} \frac{x_k}{x_k^2 + x_k} =
\sum_{k=1}^{100} \frac{x_k}{x_{k+1}}.
$$
Now consider
\[
\begin{aligned}
\frac{1}{x_{k+1}} - \frac{1}{x_k} &= \frac{x_{k+1} - x_k}{x_k x_{k+1}}\\
&= \frac{x_k^2}{x_k x_{k+1}}\\
&= \frac{x_k}{x_{k+1}}.
\end{aligned}
\]
So our sum becomes
$$
\sum_{k=1}^{100} \frac{1}{x_{k+1}} - \frac{1}{x_k} =
\frac{1}{x_{101}} - \frac{1}{x_1} = \frac{1}{x_{101}} - 2.
$$
Now the arguement should read $x_{101}$ is very large so that quantity
is just a tad more than $-2$ meaning that it's floor is indeed just $-2$.
$\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 1.9 --- Problem 6}
Prove that
$$
\sum_{n=1}^\infty 3^{n-1}\sin^3 \frac{a}{3^n} = \frac14 (a - \sin a).
$$
\end{problem}

\begin{solution}{Solution}
Using the complex definition of sine one can show that
$$
\sin^3 \frac{a}{3^n} =
\frac14 \left( 3\sin \frac{a}{3^n} - \sin \frac{a}{3^{n-1}} \right)
$$
Plugging this into the sum we have
\[
\begin{aligned}
S &= \frac14 \sum_{n=1}^\infty
\left(
3^n \sin \frac{a}{3^n} - 3^{n-1} \sin \frac{a}{3^{n-1}}
\right) \\
&= \frac14
\left(
\lim_{n\rightarrow \infty} \left[ 3^n \sin \frac{a}{3^n} \right] - \sin a
\right) \\
&= \frac14
\left(
a - \sin a
\right)
\end{aligned}
\]
where one evaluates that limit by using, for instance Taylor expansion.
$\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 2.5 --- Problem 3}
Solve the system of equations
\[
\begin{aligned}
xy + yz + zx &= 12, \\
xyz &= 2 + x + y + z
\end{aligned}
\]
in positive real numbers $x,y,z$.
\end{problem}

\begin{solution}{Solution}
It's a nice problem with inequalities. Consider
\[
\begin{aligned}
(x+y+z)^2 &= x^2 + y^2 + z^2 + 2(xy + yz + zx) \\
&\geq 3(xy + yz + zx) && \text{(Re-arrangement ineq).} \\
&= 36
\end{aligned}
\]
Now since $x,y,z > 0$ it must be the case that $x + y + z \geq 6$.
This implies that $xyz - 2 \geq 6 \iff xyz \geq 8$. Notice that the
entire problem is symmetrical in $x,y,z$.
\vspace{0.2cm}

At least one of
$x,y,z$ must be $\geq 2$ because otherwise their sum would be $< 6$.
WLOG let $x \geq 2$. Now we have
$y + z \geq 4$ and $yz \geq 4$. Again at least one of $y,z$ must be $\geq 2$
because otherwise their sum would be $ < 4$. WLOG let $y \geq 2$. But now
these two inequalties imply that $z \geq 2$ also. So we find that
$$
12 = xy + yz + zx \geq 4 + 4 + 4 = 12
$$
and so actually none of the variables can exceed 2. Thus the only solution
is $(x,y,z) = (2,2,2)$. $\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 3.2 --- Problem 5}
Prove that the equation $x^2 + y^2 + 1 = z^2$ has infinitely many integer
solutions.
\end{problem}

\begin{solution}{Solution}
Let $k \in \mathbb{N}$. Notice that
\[
\begin{aligned}
(2k)^2 + (2k^2)^2 + 1 &= 4k^4 + 4k^2 + 1 \\
&= (2k^2 + 1)^2
\end{aligned}
\]
So the tuple $(x,y,z) = (2k, 2k^2, 2k^2 + 1)$ produces a solution for
any $k \in \mathbb{N}$ and there are infinitely many such solutions.
$\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 3.2 --- Problem 6}
Let $a$ and $b$ be integers such that there exist consecutive integers
$c$ and $d$ for which $a-b = a^2 c - b^2 d$. Prove that $|a-b|$ is a
perfect square.
\end{problem}

\begin{solution}{Solution}
Let $d = c+1$. We have
\[
\begin{aligned}
a-b &= a^2 c - b^2 c - b^2 \\
b^2 &= c(a^2 - b^2) - (a-b)\\
&= (a-b)(c(a+b)-1).
\end{aligned}
\]
Consider a prime $p$ dividing $a-b$. Due to the equality above $p$ is
also a divisor of $b^2$, which only happens if $p \mid b$.
So $p$ also divides $(a-b) + 2b = a+b$. Since $c(a+b) - 1$ is coprime to
$a+b$, we have that $p \nmid [c(a+b) - 1]$. \vspace{0.2cm}

Thus we have found that $\gcd(a-b, c(a+b)-1) = 1$. Note that
$\nu_p(b^2)$ is always even. If $\nu_p(a-b)$ happened to be odd for some
prime $p$ then this means that $\nu_p(c(a+b)-1)$ must also be odd, but
this means that $a-b$ and $c(a+b)-1$ share a common factor, a contradiction.
$\blacksquare$
\end{solution}
\newpage

\begin{problem}{Chapter 1.10 --- Problem 6}
The sequence $(x_n)_n$ satisfies $\sqrt{x_{n+2} + 2} \leq x_n \leq 2$
for all $n \geq 1$. Find all possible values of $x_{1986}$.
\end{problem}

\begin{solution}{Solution}
\textit{Claim 1:} $x_n \geq 0$ for $n \in \mathbb{N}$. \vspace{0.2cm}

\textit{Proof:} Suppose $x_n < 0$. Then $\sqrt{x_{n+2} + 2} \leq x_n < 0$.
This is impossible since the square root only gives values $\geq 0$.
$\square$ \\

We can re-arrange the given inequality to $x_{n+2} + 2 \leq x_n^2$.
Since $x_{n+2} \geq 0$ it must be the case that $x_n \geq \sqrt{2}$ for
$n \in \mathbb{N}$. Define $\alpha = \sqrt2$. We now have
$$
x_n^2 \geq 2 + \alpha \ \Rightarrow \ x_n \geq \sqrt{\alpha + 2}.
$$
Define $B_{k+1} = \sqrt{B_k + 2}, B_0 = \sqrt{\alpha + 2}$, for
$k = 1,2,\ldots$. We are going to show that
$x_n \geq \lim_{k \rightarrow \infty} B_k$.
Suppose that $x_n \geq B_m$ for some non-negative integer $m$. Then
$$
B_m + 2 \leq x_n^2 \ \Rightarrow \ \sqrt{B_m + 2} = B_{m+1} \leq x_n.
$$
So for any large $N$, $x_n \geq B_N$ via induction. Now we aim to compute
$B' := \lim_{k \rightarrow \infty} B_k$.
$$
B' = \sqrt{B' + 2} \ \Rightarrow \ B' = 2
$$
via the quadratic formula and the fact that $B' > 0$. So we have deduced
that $x_n \geq 2$. But we also have $x_n \leq 2$. Thus the only sequence
that works is the constant sequence $x_n = 2$. The only value of $x_{1986}$
is 2. $\blacksquare$
\end{solution}

\end{document}
