\documentclass{article}
\usepackage{../math_template}

\author{Jonathan Kasongo}
\title{IMO Shortlist 2012 --- C1}

\begin{document}
\maketitle

\begin{problem}{IMO Shortlist 2012 --- C1}
Several positive integers are written in a row.
Iteratively, Alice chooses two adjacent numbers $x$
and $y$ such that $x > y$ and $x$ is to the left of $y$, and replaces the
pair $(x, y)$ by either $(y + 1, x)$ or
$(x - 1, x)$. Prove that she can perform only finitely many such iterations.
\end{problem}

\begin{solution}{Solution}
Let the positive integers be $a_1, a_2, \ldots, a_n$. Notice that
if $(a_n)$ becomes increasing, then Alice will no longer be able to
perform any operations.\vspace{0.2cm}


Let $x>y$ be adjacent numbers with $x$ on the left of $y$.
Since $\max(y+1, x-1) \leq x$, once we do the operation on $(x,y)$ the
number on the right will be $\geq$ the number on the left. Suppose FTSOC
that it's possible to perform an infinite number of iterations.
\vspace{0.2cm}

\textit{Claim 1: Let $M$ be the largest number such that
$a_M = \max(a_n)$.
Eventually it must end up at the end of the list.}\vspace{0.2cm}

\textit{Proof:}
Notice that the largest number in the list can only ever move right
under this operation. Suppose FTSOC that the claim is false. That is, after
a number of operations the number $a_M$ always stays in some fixed position
$k<n$ in the list. Consider the lists $a_1,a_2 \ldots, a_{k-1}$ and
$a_{k+1}, a_{k+2}, \ldots, a_n$. Now since claim 1 is false the largest
number in each list will have some fixed position that isn't the end
of the list after some number of operations also.
If we keep repeating this arguement then eventually we will reach a point
where all the numbers in $(a_n)$ have fixed positions and so no more
operations can be performed, a contradiction. $\square$ \vspace{0.2cm}

So due to claim 1, the largest number $a_M$ will end up at the end of the list
after finitely many operations. Now since it's the largest number it
can no longer move. Now consider the list $a_1, a_2, \ldots, a_{n-1}$.
Due to claim 1, the largest number in this new list will eventually end up
at the end of the list after finitely many operations also. If we repeat
this arguement then we will show that $(a_n)$ will eventually become sorted
in increasing order after finitely many operations, and so after this point
Alice will no longer be able to perform any more operations, a
contradiction.
$\blacksquare$
\end{solution}

\end{document}
